---
title: "Explore stplanr"
output: html_document
---

```{r, include=FALSE}
library(tidyverse)
library(stplanr)
library(dodgr)
library(osmdata)
library(sf)
library(leaflet)
```


## osmnx functionality
- getting data from OSM
- converting spatial network to graph data structure
- routing through the spatial graph
- adding flows

## Getting data from OSM
Unlike osmnx, which includes a set of OSM data retrieval functionality, stplanr depends on other packages to retrieve data from OSM. [osmdata](https://cran.r-project.org/web/packages/osmdata/vignettes/osmdata.html) is a simple package that can query OSM and extract data as a sf dataframe. 

```{r}
wellington_roads <- opq(bbox = 'Wellington New Zealand') %>%
  add_osm_feature(key = 'highway') %>%
  osmdata_sf()

# Extract the line geometries
wellington_roads_lines <- wellington_roads$osm_lines
```


# Plotting data from OSM
```{r}
# Static plot
# ggplot(wellington_roads_lines) + geom_sf()

# interactive plot
# osmdata gives name to geometry column
# https://github.com/ropensci/osmdata/issues/100
names(wellington_roads_lines$geometry) <- NULL

p <- leaflet(wellington_roads_lines) %>%
  addTiles() %>%
  addPolylines(label = ~name)
p
```

# Converting road network to graph structure

