---
title: "Identifying spatial network issues"
output: 
  html_document:
    keep_md: yes
---

## Spatial networks: the power and challenge
I've blogged many times previously about the insights enabled by spatial networks. But, the availability and quality of spatial network data can be challenging. I've so far utilised the vast Openstreetmap (OSM) via `osmnx` for my analyses. However, `omsnx` doesn't work well with spatial data created without the OSM taxonomy. As part of some side projects at work (and a general interest) I've wanted to work with rail network data. I've so far been able to get versions of New Zealand rail networks from a bunch of sources - from [LINZ](https://data.linz.govt.nz/search/?q=railway) to [Kiwirail open data](https://data-kiwirail.opendata.arcgis.com/datasets/13d266cb6dd141879daa76d993e2b0cc_0). After a frustrating weekend trying to understand why I couldn't route on the network, I realised that I need to amend my network analysis workflow to include pre-processing and diagnostic tools. This post goes through a simple diagnostic for checking network connectivity and highlights basic steps to create a connected, routeable network. The corrections I've applied may not be sufficient for a different network / use case but they are a great starting point. 


## Set up
All code for this post can be found here. The `renv.lock` provides the package dependencies to run this project - though it is far from a parsimonious specification as I use my projects to explore. The package can be reduced considerably as there are several package requirements (e.g. ggraph, leaflet) not needed for this particular example. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(sfnetworks)
library(tidyverse)
library(tidygraph)
library(sf)
library(osmdata)
library(mapview)
library(here)
library(rmarkdown)
```

I downloaded the New Zealand rail network from the [kiwirail open data hub](https://data-kiwirail.opendata.arcgis.com/datasets/13d266cb6dd141879daa76d993e2b0cc_0/data?geometry=103.420%2C-51.783%2C-116.834%2C-28.693) as a geodatabase but other formats are also available. 

```{r}
# kiwirail data from 
# https://data-kiwirail.opendata.arcgis.com/datasets/13d266cb6dd141879daa76d993e2b0cc_0/data?geometry=103.420%2C-51.783%2C-116.834%2C-28.693

# Basic filter and transform of rail network from Kiwirail
# only keeping tracks
nz_rail <- st_read(here::here("data", "kiwirail.gdb")) %>%
  filter(!Type %in% c("Crossover", "Yard Track")) %>%
  st_transform(2193) %>% 
  st_cast("LINESTRING")

railway_net <- nz_rail %>%
  as_sfnetwork(directed=F) 
```

```{r}
# nz port locations
port_locs <- read_csv(glue::glue("~/Documents",
                                 "infrastructure-analytics",
                                 "data",
                                 "port_locations.csv",
                                 .sep="/")) %>%
  st_as_sf(coords=c("lon", "lat"), crs=4326) %>%
  filter(country == "New Zealand") %>%
  filter(LOCODE %in% c("NZAKL", "NZTRG", "NZNPE", "NZWLG",
                       "NZNSN", "NZLYT", "NZPOE")) %>%
  st_transform(crs=2193)


```


```{r}

nodes_rail <- railway_net %>% activate("nodes") %>% st_as_sf()
orig_dest <- bind_cols(tibble(to_port = st_sfc(st_point(c(1758271, 5921462)))) %>% 
                       st_as_sf(crs = 2193), 
                       tibble(from_port = st_sfc(st_point(c(1749904, 5428126)))) %>% 
                       st_as_sf(crs = 2193)) %>%
  st_as_sf(crs = 2193) %>%
  mutate(to_index = st_nearest_feature(to_port, nodes_rail), 
         from_index = st_nearest_feature(from_port, nodes_rail),
         route = row_number())

# function that needs to be converted to pmap
st_network_paths_mod <- function(from, to){
  return(try(st_network_paths(railway_net, from, to)))
}

routes_df <- orig_dest %>%
  mutate(path = pmap(list(from = from_index, to = to_index), .f=st_network_paths_mod)) %>% 
  unnest(cols=c(path))

routes_df


```


```{r}
open_ended_nodes <- railway_net %>% 
  activate("nodes") %>% 
  mutate(degree = centrality_degree()) %>% 
  st_as_sf() %>%
  mutate(row = row_number()) %>%
  filter(degree == 1) 

disconnected_edges <- railway_net %>% 
  activate("edges") %>% 
  st_as_sf() %>% 
  filter(from %in% open_ended_nodes$row | to %in% open_ended_nodes$row)

```


```{r}

mapview(railway_net %>% activate("edges") %>% st_as_sf(), layer.name="rail network") + 
  mapview(disconnected_edges, color="red", layer.name = "nodes with only 1 edge") + 
  mapview(open_ended_nodes, color="red", col.regions="red", layer.name = "edges of 1 edge nodes")

```


## Managing the issues

```{r}

# rounding coordinates to prevent precision errors
# which created disconnected edges in the network
st_geometry(nz_rail) <- nz_rail %>% 
  st_geometry() %>% 
  map(~round(., -1)) %>% 
  st_sfc(crs = st_crs(nz_rail))

# subdividing edges where intersections happen at 
# internal points. excluding this creates a disconnected network
# because sfnetwork only connects edges where edges join at the 
# terminal nodes
railway_net <- nz_rail %>%
  as_sfnetwork(directed=F) %>% 
  convert(to_spatial_subdivision)

```



```{r}
open_ended_nodes <- railway_net %>% 
  activate("nodes") %>% 
  mutate(degree = centrality_degree()) %>% 
  st_as_sf() %>%
  mutate(row = row_number()) %>%
  filter(degree == 1) 

disconnected_edges <- railway_net %>% 
  activate("edges") %>% 
  st_as_sf() %>% 
  filter(from %in% open_ended_nodes$row | to %in% open_ended_nodes$row)

```


```{r}

mapview(railway_net %>% activate("edges") %>% st_as_sf(), layer.name="rail network") + 
  mapview(disconnected_edges, color="red", layer.name = "nodes with only 1 edge") + 
  mapview(open_ended_nodes, color="red", col.regions="red", layer.name = "edges of 1 edge nodes")

```
