---
title: "Identifying spatial network issues"
output: 
  html_document:
    keep_md: yes
---

## Spatial networks: the power and challenge
I've blogged many times previously about the insights enabled by spatial networks. But, the availability and quality of spatial network data can be challenging. I've so far utilised the vast Openstreetmap (OSM) via `osmnx` for my analyses. However, `omsnx` doesn't work well with spatial data created without the OSM taxonomy. As part of some side projects at work (and a general interest) I've wanted to work with rail network data. I've so far been able to get versions of New Zealand rail networks from a bunch of sources - from [LINZ](https://data.linz.govt.nz/search/?q=railway) to [Kiwirail open data](https://data-kiwirail.opendata.arcgis.com/datasets/13d266cb6dd141879daa76d993e2b0cc_0). After a frustrating weekend trying to understand why I couldn't route on the network, I realised that I need to amend my network analysis workflow to include pre-processing and diagnostic tools. This post goes through a simple diagnostic for checking network connectivity and highlights basic steps to create a connected, routeable network. The corrections I've applied may not be sufficient for a different network / use case but they are a great starting point. 


## Set up
All code for this post can be found here. The `renv.lock` provides the package dependencies to run this project - though it is far from a parsimonious specification as I use my projects to explore. The package can be reduced considerably as there are several package requirements (e.g. ggraph, leaflet) not needed for this particular example. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(sfnetworks)
library(tidyverse)
library(tidygraph)
library(sf)
library(osmdata)
library(mapview)
library(here)
library(rmarkdown)
```

I downloaded the New Zealand rail network from the [kiwirail open data hub](https://data-kiwirail.opendata.arcgis.com/datasets/13d266cb6dd141879daa76d993e2b0cc_0/data?geometry=103.420%2C-51.783%2C-116.834%2C-28.693) as a geodatabase but other formats are also available. 

```{r}
# kiwirail data from 
# https://data-kiwirail.opendata.arcgis.com/datasets/13d266cb6dd141879daa76d993e2b0cc_0/data?geometry=103.420%2C-51.783%2C-116.834%2C-28.693

# Basic filter and transform of rail network from Kiwirail
# only keeping tracks
nz_rail <- st_read(here::here("data", "kiwirail.gdb")) %>%
  filter(!Type %in% c("Crossover", "Yard Track")) %>%
  st_transform(2193) %>% 
  st_cast("LINESTRING")
```

Included in this repo is also a file of port locations from around the world. Though this example will only use a subset of New Zealand ports. 
```{r}
# nz port locations
port_locs <- read_csv(here::here("data",
                                 "port_locations.csv")) %>%
  st_as_sf(coords=c("lon", "lat"), crs=4326) %>%
  filter(country == "New Zealand") %>%
  filter(LOCODE %in% c("NZAKL", "NZTRG", "NZNPE", "NZWLG",
                       "NZNSN", "NZLYT", "NZPOE")) %>%
  st_transform(crs=2193)


```


## Creating a routeable rail network
The excellent new `sfnetworks` package offers a very simple way to create a routeable network, a graph structure, from any spatial points or lines dataset. Note, the railway network, originally a `multi-linestring` spatial lines dataset has to be cast as a simpler `linestring` for the conversion to an `sfnetwork` data format. This is done at the outset - when the dataset is imported. 

```{r}
railway_net <- nz_rail %>%
  as_sfnetwork(directed=F) 
```


## Associating external data with network nodes
In many cases of routing problems, we're interested in the route to a point of interest (POI) that is defined separately. Here, the POIs are port locations. For routing to be possible, these POIs need to be associated to a node in the `sfnetwork` graph. For the inclusion of new The new `st_nearest_feature` is a helper function in the `sfnetworks` package that finds the index of the closest feature (nodes or edges depending on choice) to the POI. In this example, we are trying to find a route from Auckland to Wellington, New Zealand. 

```{r}

nodes_rail <- railway_net %>% activate("nodes") %>% st_as_sf()

from = "Auckland"
to = "Wellington"


orig_dest <- bind_cols(tibble(from_port = port_locs %>% filter(port_name == from) %>% pull(geometry)), 
                       tibble(to_port = port_locs %>% filter(port_name == to) %>% pull(geometry))) %>%
  st_as_sf(crs = 2193) %>%
  mutate(to_index = st_nearest_feature(to_port, nodes_rail), 
         from_index = st_nearest_feature(from_port, nodes_rail),
         route = row_number())
```

## Routing
Once the indices of the POIs are found, we can use the `st_network_paths` wrapper function to find the shortest path between a single point of origin and a vector of destinations. However, the function returns no path as seen by the empty list in the `edge_paths` column. 

```{r}
# function that needs to be converted to pmap
st_network_paths_mod <- function(from, to){
  return(try(st_network_paths(railway_net, from, to)))
}

routes_df <- orig_dest %>%
  mutate(path = pmap(list(from = from_index, to = to_index), .f=st_network_paths_mod)) %>% 
  unnest(cols=c(path))

routes_df


```

## Diagnosing issues
The difficulty with spatial networks is that they may appear connected to the naked eye but there are some insidious issues that manifest in a disconnected graph. A simple diagnostic is to examine the connectivity of all the nodes in the network. A good metric is the node `degree` - the number of edges connected to any node. When the degree is `1`, the node is only connected to one edge. Terminal nodes, where the tracks end, are legitimate single degree nodes. However, as the interactive graph below shows, the whole network is comprised of disconnected single connection nodes. No wonder the routing algorithm couldn't find a path! 

```{r}
open_ended_nodes <- railway_net %>% 
  activate("nodes") %>% 
  mutate(degree = centrality_degree()) %>% 
  st_as_sf() %>%
  mutate(row = row_number()) %>%
  filter(degree == 1) 

disconnected_edges <- railway_net %>% 
  activate("edges") %>% 
  st_as_sf() %>% 
  filter(from %in% open_ended_nodes$row | to %in% open_ended_nodes$row)


# Visualise connectivity
mapview(railway_net %>% activate("edges") %>% st_as_sf(), layer.name="rail network") + 
  mapview(disconnected_edges, color="red", layer.name = "nodes with only 1 edge") + 
  mapview(open_ended_nodes, color="red", col.regions="red", layer.name = "edges of 1 edge nodes")

```


## Managing the issues

In my example, there were two key issues preventing a connected network: (1) too high precision of coordinates leading to small gaps between what should be connected edges and, (2) edges connected to interior nodes. The first is a data problem. The degree of rounding to the nearest 10 m is possibly a little high, but examining some of the disconnected areas identified gaps of a few metres. The second problem appears to be a peculiarity of the `sfnetworks` paradigm where [edges that aren't connected at terminal nodes are considered disconnected](https://luukvdmeer.github.io/sfnetworks/articles/preprocess_and_clean.html#subdivide-edges-1). Luckily solving these two issues is not too challenging with sensible suggestions [here](https://gis.stackexchange.com/questions/370640/how-to-connect-edges-in-a-network-even-if-they-dont-exactly-match-spatially) for the first problem and `sfnetworks` documentation on the operation `to_spatial_subdivision` for the second. 

```{r}

# rounding coordinates to prevent precision errors
# which created disconnected edges in the network
st_geometry(nz_rail) <- nz_rail %>% 
  st_geometry() %>% 
  map(~round(., -1)) %>% 
  st_sfc(crs = st_crs(nz_rail))

# subdividing edges where intersections happen at 
# internal points. excluding this creates a disconnected network
# because sfnetwork only connects edges where edges join at the 
# terminal nodes
railway_net <- nz_rail %>%
  as_sfnetwork(directed=F) %>% 
  convert(to_spatial_subdivision)

```


## A connected network?

If we repeat the same routing calculation between the two points, we see now that there is a path!

```{r}

nodes_rail <- railway_net %>% activate("nodes") %>% st_as_sf()

orig_dest <- bind_cols(tibble(from_port = port_locs %>% filter(port_name == from) %>% pull(geometry)), 
                       tibble(to_port = port_locs %>% filter(port_name == to) %>% pull(geometry))) %>%
  st_as_sf(crs = 2193) %>%
  mutate(to_index = st_nearest_feature(to_port, nodes_rail), 
         from_index = st_nearest_feature(from_port, nodes_rail),
         route = row_number())

routes_df <- orig_dest %>%
  mutate(path = pmap(list(from = from_index, to = to_index), .f=st_network_paths_mod)) %>% 
  unnest(cols=c(path))

auck_wlg <- routes_df %>% 
  unnest(edge_paths) %>% 
  select(edge_paths) %>% 
  st_drop_geometry() %>% 
  inner_join(railway_net %>% 
               activate("edges") %>% 
               st_as_sf() %>% 
               mutate(edge = row_number()), by = c("edge_paths" = "edge"))  %>% 
  st_as_sf()

# Visualise route
ggplot() + 
  geom_sf(data = railway_net %>% activate("edges") %>% st_as_sf()) + 
  geom_sf(data=auck_wlg, colour="red") 
```


## Why does routing work now?

Repeating the calculation of the node degree, we now see that there are very few disconnected nodes in the corrected network. The nodes that are disconnected and not at the end of the tracks are actually odd little side tracks - probably around railway stations and junctions. 

```{r}
open_ended_nodes <- railway_net %>% 
  activate("nodes") %>% 
  mutate(degree = centrality_degree()) %>% 
  st_as_sf() %>%
  mutate(row = row_number()) %>%
  filter(degree == 1) 

disconnected_edges <- railway_net %>% 
  activate("edges") %>% 
  st_as_sf() %>% 
  filter(from %in% open_ended_nodes$row | to %in% open_ended_nodes$row)

# Visualise corrected network
mapview(railway_net %>% activate("edges") %>% st_as_sf(), layer.name="rail network") + 
  mapview(disconnected_edges, color="red", layer.name = "nodes with only 1 edge") + 
  mapview(open_ended_nodes, color="red", col.regions="red", layer.name = "edges of 1 edge nodes")

```
